---
title: 'Guided project: Baseball'
author: "Fredrick Boshe"
date: "21/04/2021"
output: 
  github_document: default
  rmarkdown::github_document: default
---

```{r setup, include=FALSE, results='hide', message=FALSE, warning=FALSE}

#Loading packages
library(tidyverse)
library(DBI)
library(RSQLite)
library(ggplot2)
library(dplyr)
library(corrplot)
library(plotly)
library(ggcorrplot)
library(scales)
library(tufte)
library(lubridate)
library(streamgraph)
library(jsonlite)
library(emo)
library(zoo)


```
It looks like the game log has a record of over 170,000 games. It looks like these games are chronologically ordered and occur between 1871 and 2016.

For each game we have:

* general information on the game
* team level stats for each team
* a list of players from each team, numbered, with their defensive positions
* the umpires that officiated the game
* some 'awards', like winning and losing pitcher

We have a game_log_fields.txt file that tell us that the player number corresponds with the order in which they batted.

It's worth noting that there is no natural primary key column for this table.

```{r data, include=TRUE, message=FALSE, warning=FALSE}
game<-read_csv("C:/Users/ricky/OneDrive/Desktop/School/R/DataQuest/SQL/Guided project/game_log.csv")

park<-read_csv("C:/Users/ricky/OneDrive/Desktop/School/R/DataQuest/SQL/Guided project/park_codes.csv")

people<-read_csv("C:/Users/ricky/OneDrive/Desktop/School/R/DataQuest/SQL/Guided project/person_codes.csv")

team<-read_csv("C:/Users/ricky/OneDrive/Desktop/School/R/DataQuest/SQL/Guided project/team_codes.csv")

##Load the tables into a database
conn <- dbConnect(SQLite(), "mlb.db") #initialize a new database

dbWriteTable(conn=conn, name = "game", value = game, row.names=FALSE, 
             header=TRUE)
dbWriteTable(conn=conn, name = "park", value = park, row.names=FALSE, 
             header=TRUE)
dbWriteTable(conn=conn, name = "people", value = people, row.names=FALSE, 
             header=TRUE)
dbWriteTable(conn=conn, name = "team", value = team, row.names=FALSE, 
             header=TRUE)

#Check if all tables have been added to the database
tables <- dbListTables(conn)
tables

```
## Defensive Positions

In the game log, each player has a defensive position listed, which seems to be a number between 1-10. Doing some research around this, I found this [article] (http://probaseballinsider.com/baseball-instruction/baseball-basics/baseball-basics-positions/) which gives us a list of names for each numbered position:

* Pitcher
* Catcher
* 1st Base
* 2nd Base
* 3rd Base
* Shortstop
* Left Field
* Center Field
* Right Field

The 10th position isn't included, it may be a way of describing a designated hitter that does not field. I can find a retrosheet page that indicates that position 0 is used for this, but we don't have any position 0 in our data. I have chosen to make this an 'Unknown Position' so I'm not including data based on a hunch.

## Leagues

Wikipedia tells us there are currently two leagues - the American (AL) and National (NL). Upon investigation of the data, we see that there are actually 4 more. After some googling, we come up with:

* NL: National League
* AL: American League
* AA: [American Association](https://en.wikipedia.org/wiki/American_Association_%2819th_century%29)
* FL: [Federal League](https://en.wikipedia.org/wiki/Federal_League)
* PL: [Players League](https://en.wikipedia.org/wiki/Players%27_League)
* UA: [Union Association](https://en.wikipedia.org/wiki/Union_Association)

It also looks like we have about 1000 games where the home team doesn't have a value for league.

```{sql, connection=conn}
--Preview the tables in the database
SELECT
    name,
    type
FROM sqlite_master
WHERE type IN ("table","view");

```


```{sql, connection=conn}
--Create a new column in the game table
ALTER TABLE game
ADD COLUMN game_id TEXT;

```


```{sql, connection=conn}
--Create fill in rows
UPDATE game
SET game_id = h_name||date||number_of_game;

```

```{sql, connection=conn}
--Check if query worked
SELECT game_id, date,
       h_name, number_of_game
FROM game
LIMIT 5;

```

# Looking for Normalization Opportunities

The following are opportunities for normalization of our data:

* In person_codes, all the debut dates will be able to be reproduced using game log data.
* In team_codes, the start, end and sequence columns will be able to be reproduced using game log data.
* In park_codes, the start and end years will be able to be reproduced using game log data. While technically the state is an attribute of the city, we might not want to have a an incomplete city/state table so we will leave this in.
* There are lots of places in game log where we have a player ID followed by the players name. We will be able to remove this and use the name data in person_codes
* In game_log, all offensive and defensive stats are repeated for the home team and the visiting team. We could break these out and have a table that lists each game twice, one for each team, and cut out this column repetition.
* Similarly, in game_log, we have a listing for 9 players on each team with their positions - we can remove these and have one table that tracks player appearances and their positions.
* We can do a similar thing with the umpires from game_log, instead of listing all four positions as columns, we can put the umpires either in their own table or make one table for players, umpires and managers.
* We have several awards in game_log like winning pitcher and losing pitcher. We can either break these out into their own table, have a table for awards, or combine the awards in with general appearances like the players and umpires.


##Database Schema




```{sql, connection=conn}
--Create Person table
CREATE TABLE person (person_id TEXT PRIMARY KEY,
              first_name TEXT,
              second_name TEXT);
```

```{sql, connection=conn}
--Populate person table
INSERT INTO person
SELECT p.id, p.first, p.last 
FROM people as p; 

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM person
LIMIT 5;

```

```{sql, connection=conn}
--Create park table
CREATE TABLE park_new (park_id TEXT PRIMARY KEY,
                   name TEXT,
                   nickname TEXT,
                   city TEXT,
                   state TEXT,
                   notes TEXT);
 

```

```{sql, connection=conn}
--Populate park table
INSERT INTO park_new
SELECT pk.park_id, pk.name, pk.aka, pk.city, pk.state, pk.notes 
FROM park as pk; 

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM park_new
LIMIT 5;

```

```{sql, connection=conn}
--Create league table
CREATE TABLE IF NOT EXISTS league (
      league_id TEXT PRIMARY KEY,
      name TEXT); -- If NOT EXIST helps to ensure no duplicates

```

Insert league values i.e. names of the six league IDs. (not present in any existing table so far)

```{sql, connection=conn}
--Populate league table
INSERT INTO league
VALUES("NL", "National League"),
      ("AL", "American League"),
      ("AA", "American Association"),
      ("FL", "Federal League"),
      ("PL", "Players League"),
      ("UA", "Union Association"); 

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM league
LIMIT 5;

```

The data for the table "appearance_type" exists in a separate csv file. Import and load values to the table.

```{sql, connection=conn}
--Make sure no pre-existing appearance_type table
DROP TABLE IF EXISTS appearance_type; 

```

```{r appearance, include= TRUE}

appearance_type<-read_csv("C:/Users/ricky/OneDrive/Desktop/School/R/DataQuest/SQL/Guided project/appearance_type.csv")

##Load the table into the database
dbWriteTable(conn=conn, name = "appearance_type", value = appearance_type, 
             row.names=FALSE, header=TRUE)
duplicated(team$team_id)

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM appearance_type
LIMIT 5;

```


```{sql, connection=conn}
--Create team_new table
CREATE TABLE team_new (team_id TEXT PRIMARY KEY,
                       league_id TEXT,
                       city TEXT,
                       nickname TEXT,
                       franch_id,
                       FOREIGN KEY (league_id) 
                       REFERENCES league(league_id));
 
```

```{sql, connection=conn}
--Populate team_new table
INSERT OR IGNORE INTO team_new --Insert or Ignore helps deal with any duplicates on Primary Key
SELECT t.team_id, t.league, t.city, t.nickname, t.franch_id 
FROM team as t; 

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM team_new
LIMIT 5;

```


```{sql, connection=conn}
--Create game_new table
CREATE TABLE game_new (game_id TEXT PRIMARY KEY,
                       date INTEGER,
                       number_of_game INTEGER,
                       park_id TEXT,
                       lengths_out INTEGER,
                       day BOOLEAN,
                       completion TEXT,
                       forfeit TEXT,
                       protest TEXT,
                       attendance NUMERIC,
                       lengths_minute NUMERIC,
                       additional_info TEXT,
                       acquisition_info TEXT,
                       FOREIGN KEY (park_id) 
                       REFERENCES park_new(park_id));
 
```

sqlite does not have the "BOOLEAN" data type. We must code the columns to 1 or 0 before passing them to the table we are creating. 
```{sql, connection=conn}
--Populate game_new table
INSERT OR IGNORE INTO game_new 
SELECT game_id, date, number_of_game, park_id,length_outs,
CASE
       WHEN day_night = "D" THEN 1
       WHEN day_night = "N" THEN 0
       ELSE NULL
       END
       AS day,
      completion, forfeit, protest, attendance,
      length_minutes, additional_info, acquisition_info
FROM game; 

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM game_new
LIMIT 5;

```

```{sql, connection=conn}
--Create team_appearance table
CREATE TABLE IF NOT EXISTS team_appearance (
      team_id TEXT,
      game_id TEXT,
      home BOOLEAN,
      league_id TEXT,
      score INTEGER,
      line_score TEXT,
      at_bats INTEGER,
      hits INTEGER,
      doubles INTEGER,
      triples INTEGER,
      homeruns INTEGER,
      rbi INTEGER,
      sacrifice_hits INTEGER,
      sacrifice_flies INTEGER,
      hit_by_pitch INTEGER,
      walks INTEGER,
      intentional_walks INTEGER,
      strikeouts INTEGER,
      stolen_bases INTEGER,
      caught_stealing INTEGER,
      grounded_into_double INTEGER,
      first_catcher_interference INTEGER,
      left_on_base INTEGER,
      pitchers_used INTEGER,
      individual_earned_runs INTEGER,
      team_earned_runs INTEGER,
      wild_pitches INTEGER,
      balks INTEGER,
      putouts INTEGER,
      assists INTEGER,
      errors INTEGER,
      passed_balls INTEGER,
      double_plays INTEGER,
      triple_plays INTEGER,
      PRIMARY KEY (team_id, game_id),
      FOREIGN KEY (team_id) REFERENCES team_new(team_id),
      FOREIGN KEY (game_id) REFERENCES game_new(game_id),
      FOREIGN KEY (team_id) REFERENCES team_new(team_id)
  );

 
```

sqlite does not have the "BOOLEAN" data type. We must code the columns to 1 or 0 before passing them to the table we are creating. 
```{sql, connection=conn}
--Populate team_appearance table
INSERT OR IGNORE INTO team_appearance
      SELECT
          h_name,
          game_id,
          1 AS home,
          h_league,
          h_score,
          h_line_score,
          h_at_bats,
          h_hits,
          h_doubles,
          h_triples,
          h_homeruns,
          h_rbi,
          h_sacrifice_hits,
          h_sacrifice_flies,
          h_hit_by_pitch,
          h_walks,
          h_intentional_walks,
          h_strikeouts,
          h_stolen_bases,
          h_caught_stealing,
          h_grounded_into_double,
          h_first_catcher_interference,
          h_left_on_base,
          h_pitchers_used,
          h_individual_earned_runs,
          h_team_earned_runs,
          h_wild_pitches,
          h_balks,
          h_putouts,
          h_assists,
          h_errors,
          h_passed_balls,
          h_double_plays,
          h_triple_plays
      FROM game
  
  UNION
  
      SELECT    
          v_name,
          game_id,
          0 AS home,
          v_league,
          v_score,
          v_line_score,
          v_at_bats,
          v_hits,
          v_doubles,
          v_triples,
          v_homeruns,
          v_rbi,
          v_sacrifice_hits,
          v_sacrifice_flies,
          v_hit_by_pitch,
          v_walks,
          v_intentional_walks,
          v_strikeouts,
          v_stolen_bases,
          v_caught_stealing,
          v_grounded_into_double,
          v_first_catcher_interference,
          v_left_on_base,
          v_pitchers_used,
          v_individual_earned_runs,
          v_team_earned_runs,
          v_wild_pitches,
          v_balks,
          v_putouts,
          v_assists,
          v_errors,
          v_passed_balls,
          v_double_plays,
          v_triple_plays
      from game; 

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM team_appearance
LIMIT 5;

```


```{sql, connection=conn}
--Create person_appearance table
CREATE TABLE IF NOT EXISTS person_appearance (
      appearance_id TEXT PRIMARY KEY,
      person_id TEXT,
      team_id TEXT,
      game_id TEXT,
      appearance_type_id TEXT,
      FOREIGN KEY (person_id) REFERENCES person(person_id),
      FOREIGN KEY (team_id) REFERENCES team_new(team_id),
      FOREIGN KEY (game_id) REFERENCES game_new(game_id),
      FOREIGN KEY (appearance_type_id) REFERENCES appearance_type(appearance_type_id));
 
```


```{sql, connection=conn}
--Populate person_appearance table
INSERT OR IGNORE INTO person_appearance (
      game_id,
      team_id,
      person_id,
      appearance_type_id
  ) 
      SELECT
          game_id,
          NULL,
          hp_umpire_id,
          "UHP"
      FROM game
      WHERE hp_umpire_id IS NOT NULL    
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [1b_umpire_id],
          "U1B"
      FROM game
      WHERE "1b_umpire_id" IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [2b_umpire_id],
          "U2B"
      FROM game
      WHERE [2b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          [3b_umpire_id],
          "U3B"
      FROM game
      WHERE [3b_umpire_id] IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          lf_umpire_id,
          "ULF"
      FROM game
      WHERE lf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          NULL,
          rf_umpire_id,
          "URF"
      FROM game
      WHERE rf_umpire_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_manager_id,
          "MM"
      FROM game
      WHERE v_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_manager_id,
          "MM"
      FROM game
      WHERE h_manager_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_pitcher_id,
          "AWP"
      FROM game
      WHERE winning_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score < v_score THEN h_name
              ELSE v_name
              END,
          losing_pitcher_id,
          "ALP"
      FROM game
      WHERE losing_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          saving_pitcher_id,
          "ASP"
      FROM game
      WHERE saving_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          CASE
              WHEN h_score > v_score THEN h_name
              ELSE v_name
              END,
          winning_rbi_batter_id,
          "AWB"
      FROM game
      WHERE winning_rbi_batter_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          v_name,
          v_starting_pitcher_id,
          "PSP"
      FROM game
      WHERE v_starting_pitcher_id IS NOT NULL
  
  UNION
  
      SELECT
          game_id,
          h_name,
          h_starting_pitcher_id,
          "PSP"
      FROM game
      WHERE h_starting_pitcher_id IS NOT NULL;
 

```

```{sql, connection=conn}
--Check if query worked
SELECT *
FROM person_appearance
LIMIT 5;

```

```{sql, connection=conn}
--Drop orriginal unnormalized tables
SELECT *
FROM person_appearance
LIMIT 5;

```



```{r disconnect, include=FALSE}
dbListTables(conn)

#remove original un-normalized tables
tables <- c("game", "park",
            "team", "people")
for (t in tables) {
  drop_command = sprintf("DROP TABLE %s", t)
  dbExecute(conn, drop_command)
}

dbListTables(conn)
dbDisconnect(conn)

```
